---
title: "Birth Weights Code"
author: "Aaron Oustrich, Josh Bergstrom"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
# link https://mheaton.byu.edu/docs/files/Stat469/IID/Stat469IIDClassRCode.html

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(ggplot2)
library(GGally)
library(car)
library(MASS)
library(lmtest)

weights <- read_table("data/BirthWeights.txt")

# Convert Race and Gen cols to factors
weights <- weights %>% 
  mutate_at(vars(Race,Gen),factor)
```

# Exploratory Data Analysis

### 1. Scatterplot of BirthWeight by Mage
```{r}
ggplot(data=weights, mapping=aes(x=BirthWeight,y=Mage)) + geom_point()
```

### 2. Side-by-side boxplots of BirthWeight for each category in Race
```{r}
ggplot(data=weights, mapping=aes(x=Race,y=BirthWeight)) + geom_boxplot()
```

### 3. A scatterplot of BirthWeight by Gage where the dots are colored according to Gen
```{r}
ggplot(data=weights, mapping=aes(x=BirthWeight,y=Gage,color=Gen)) + geom_point()
```

### 4. The correlation between BirthWeight and Mage
```{r}
cor(weights$BirthWeight,weights$Mage)
```

### 5. A pairs plot of all the variables in the BirthWeight dataset
```{r}
new_weights <- weights %>% select(-BirthWeight,BirthWeight)
ggpairs(new_weights, progress = FALSE)
```

# Fitting a Linear Model

### 1. Without the use of lm() calculate β̂ and s2. Verify your answer using lm()
(y−Xβ̂ )′(y−Xβ̂ )/(n−P−1)
```{r}
birth.lm <- lm(BirthWeight ~ ., data=weights)
X <- model.matrix(BirthWeight ~ ., data=weights)
y <- weights$BirthWeight

beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
s2 <- t(y - X %*% beta_hat) %*% (y - X %*% beta_hat) / ( nrow(weights) - ncol(X) )

sigma(birth.lm)
print(sqrt(s2))

coef(birth.lm)
print(beta_hat)
```
They are the same values.

### 2. Without the use of lm() calculate the fitted values Xβ̂ Verify your calculations by pulling off the fitted values from an lm() object.

```{r}
# Round all the fitted values to 3 decimal places
matrix_fitted_vals = round(X %*% beta_hat,3) 
lm_fitted_vals = round(fitted(birth.lm),3)

# Check if all the rounded fitted values are the same
all(matrix_fitted_vals == lm_fitted_vals)
```

### 3. Without the use of lm() calculate the residuals y−Xβ̂ Verify your calculations by pulling off the residuals from an lm() object.
```{r}
# Round all the residuals to 3 decimal places
matrix_residuals = round(y - X %*% beta_hat,3)
lm_residuals = round(residuals(birth.lm),3)

# Check if all the rounded residuals are the same
all(matrix_residuals == lm_residuals)
```

### 4. Identify your model R2 from the summary() output.
```{r}
summary(birth.lm)
```
The model R2 is 0.6065

# Checking Assumptions

### 1. Construct added variable plots and assess if the linearity assumption is OK for this data.
```{r}
avPlots(birth.lm)
```
The linearity assumption is not OK for this data. The added variable plots show that the relationship between the response and the predictor is linear.

### 2. Construct a histogram of the standardized residuals and run a KS-test to see if the normality assumption is OK for this data.
```{r}
standardized.residuals <- stdres(birth.lm)
ggplot()+geom_histogram(mapping=aes(x=standardized.residuals))
ks.test(standardized.residuals, "pnorm")
```
The Normality assumption is OK for this data. The p-value of the KS-test is 0.4884, so we confirm normality.

### 3. Draw a scatterplot of the fitted values vs. standardized residuals and run a BP-test to see if the equal variance assumption is OK for this data.
```{r}
ggplot(mapping=aes(x=fitted(birth.lm),y=standardized.residuals)) + 
  geom_point()
bptest(birth.lm)
```
The equal variance assumption is OK for this data. The residuals v. fitted values plot is a nice cloud shape. The p-value of the BP-test is 0.338, so we confirm equal variance.

# Predictions

